---
title: "Coursera 'Reproductible Research'- Assignment #1"
author: "Abdel612"
date: "May, 11th of 2015"
---

This is the R Markdown document as required by the Coursera "Reproductible Research" course's assignement #1.  



# Helper Functions
Let's start with some helper functions that allow us to uncrompress the provided data file , reading and filling the dataset up with measured values, and ending with a function computing the total number of step done in a day..

### Uncompress the provided data file (unziping the .zip file)
This function does not return anything, but makes the job to unzip the archive given by parameter *fileName*:
```{r, echo=TRUE}
#
# Unzip data archive file. Default file name="activity.zip"
#
uncompressDataArchive = function(fileName="activity.zip"){ 
    unzip(fileName, files = NULL, list = FALSE, overwrite = TRUE,
          junkpaths = FALSE, exdir = ".", unzip = "internal",
          setTimes = FALSE)
}
```

### Read the data file and load values in the returned data frame
Load any CSV formatted file given by parameter *fromFileName*

Returns a data frame corresponding to the CSV file
```{r, echo=TRUE}
#
# Load dataset according to a file name which defaulted to "activity.csv"
#
loadDataset = function(fromFileName="activity.csv"){
  dataSet=read.csv(fromFileName)
  return(dataSet)
}
```

### Compute the total number of steps by date. 
Returns a data frame made of 2 variables: 

- date : POSIX : the considered date
- steps: numeric : Total number of steps done the day date
```{r, echo=TRUE}
#
# Compute mean steps by date
#
# Return a new data.frame with 2 columns:
#     - date : POSIX Date in YYYY-MM-DD format
#     - steps: numeric : Total number of steps in the day "date"
getNumberOfStepsByDate = function(dataSet){
  totalStepPerDay = tapply(dataSet$steps, dataSet$date, sum, na.rm = TRUE)
  theDate = names(totalStepPerDay)
  theTotal = as.numeric(totalStepPerDay)
  result = data.frame(date=strptime(theDate, format="%Y-%m-%d"), steps=theTotal)
  return( result )
}
```



# Here we go with assignment questions

## Loading and preprocessing the data
```{r, echo=TRUE}
#
# PREREQUISITE: Loading and preprocessing the data
#

#   Step 1: explode the zip archive
uncompressDataArchive()

#   Step 2: Read the original dataset
originalDF = loadDataset()
```


## QUESTION 1: What is mean total number of steps taken per day?
  
###   Step 1: Calculate the total number of steps taken per day
```{r, echo=TRUE}
question1DF = getNumberOfStepsByDate( subset(originalDF, !is.na(originalDF$date)) )
```
```{r, echo=FALSE}
question1DF
```

###   Step 2: Make a histogram of the total number of steps taken each day
```{r, echo=FALSE}
hist(question1DF$steps, xlab = "5' interval", ylab = "# steps", main = "Steps distribution")
```

###   Step 3: Calculate and report the mean and median of the total number of steps taken per day

- Mean:
```{r, echo=TRUE}
meanQuestion1 = mean(question1DF$steps)
```
```{r, echo=FALSE}
meanQuestion1
```
- Median:
```{r, echo=TRUE}
medianQuestion1 = median(question1DF$steps)
```
```{r, echo=FALSE}
medianQuestion1
```

Let's save some memory by removing unused varaibels:
```{r, echo=TRUE}
rm(question1DF) # gain some memory space by removing this dataset (no more use)
```



## QUESTION 2: What is the average daily activity pattern?

###   Step 1: Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
question2List=tapply(originalDF$steps, originalDF$interval, mean, na.rm = TRUE)
plot(names(question2List), as.numeric(question2List), 
     type = "l",
     xlab = "Interval", ylab = "# daily steps", main = "Average steps")
```

###   Step 2: Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
idxMaxSteps = which.max(question2List)
intervalMaxStep = as.numeric(names(question2List)[idxMaxSteps])
intervalMaxStep.hour = floor(intervalMaxStep/60)
intervalMaxStep.minute = intervalMaxStep%%60
sprintf("5' Interval #%d (starting at %02d:%02d:00) contains the maximum average number of steps over all days in the dataset: %0.2f",
        intervalMaxStep, intervalMaxStep.hour, intervalMaxStep.minute, question2List[idxMaxSteps])
```


## QUESION 3: Imputing missing values

**Note:** There are a number of days/intervals where there are missing values (coded as NA). 
The presence of missing days may introduce bias into some calculations or summaries of the data.

###   Step 1: Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo=TRUE}
missingStepIdx = which(is.na(originalDF$steps))
sprintf("There are %d missing values in the original dataset", length(missingStepIdx) )
```

###   Step 2 & 3: Devise a strategy for filling in all of the missing values in the dataset. 
<br>The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
<br><br>               Create a new dataset that is equal to the original dataset but with the missing data filled in.
<br>
*The choosen strategy is to assign missing values the previously calculated mean (rounded) that corresponds to the same time interval*
```{r, echo=TRUE}
question3DF = originalDF
question3DF$steps[missingStepIdx] = floor( as.numeric(question2List[as.character(originalDF$interval[missingStepIdx])]) + 0.5 )
question3DF = getNumberOfStepsByDate( question3DF )
```
```{r, echo=FALSE}
question3DF
```

###   Step 4: Made of 4 steps: a, b, c and d
4.a: Make a histogram of the total number of steps taken each day
```{r, echo=FALSE}
hist(question3DF$steps, xlab = "5' interval", ylab = "# steps", main = "Steps distribution")
```

4.b: Calculate and report the mean and median total number of steps taken per day
-Mean:
```{r, echo=TRUE}
meanQuestion3 = mean(question3DF$steps)
```
```{r, echo=FALSE}
meanQuestion3
```
-Median:
```{r, echo=TRUE}
medianQuestion3 = median(question3DF$steps)
```
```{r, echo=FALSE}
medianQuestion3
```


4.c: Do these values differ from the estimates from the first part of the assignment?

**Response:** Only the mean is changing since according to the choosen strategy, the distribution does not change, thus the median won't change

4.d: What is the impact of imputing missing data on the estimates of the total daily number of steps?

**Response:** The impact of imputing missing data with ur strategy did get the mean value raising


Let's gain some more memory by removing new no more used variables
```{r, echo=TRUE}
rm(question3DF) # gain some memory space by removing this dataset (no more use)
```

## QUESION 4: Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
WEEK_DAY = "Week Day" # Constant for weekday.type factor column (see below) => It's on week day
WEEK_END = "Week-End" # Constant for weekday.type factor column (see below) => It's on week-end
question4DF = originalDF
question4DF$steps[missingStepIdx] = floor( as.numeric(question2List[as.character(originalDF$interval[missingStepIdx])]) + 0.5 )
```

###   Step 1: Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE}
question4DF$weekday = strptime(question4DF$date, format = "%Y-%m-%d")
question4DF$weekday = lapply( question4DF$weekday$wday, # Avoid localisation issue (I'm using a french locale)
                              function(x){ 
                                if (x>5)
                                  return(WEEK_END)
                                else 
                                  return(WEEK_DAY)
                                }
                              )
question4DF$weekday = as.factor(as.character(question4DF$weekday))
```

### Step 2: Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). <br> See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r, echo=TRUE}
library(lattice) # Just to get sure the package is loaded

# Build weekday dataset with total steps an interval over all considered week days 
weekdayDF = subset(question4DF, question4DF$weekday == WEEK_DAY)
weekdayAvg = tapply(weekdayDF$steps, weekdayDF$interval, mean)
weekdayAvgDF = data.frame( interval = as.numeric( names(weekdayAvg) ), steps = floor(as.numeric(weekdayAvg)+0.5), weekday.type = rep(WEEK_DAY,length(weekdayAvg) ) )

# Build weekend dataset with total steps an interval over all considered week-end day
weekendDF = subset(question4DF, question4DF$weekday == WEEK_END)
weekendAvg = tapply(weekendDF$steps, weekendDF$interval, mean)
weekendAvgDF = data.frame( interval = as.numeric( names(weekendAvg) ), steps = floor(as.numeric(weekendAvg)+0.5), weekday.type = rep(WEEK_END,length(weekendAvg) ) ) 

# The final dataset is made of the two above dataset (weekday and weekden)
resultPanel = rbind(weekdayAvgDF,weekendAvgDF)

# Plot the panel as expected by the assignment
xyplot( steps ~ interval | weekday.type, data = resultPanel, 
        layout=c(1,2), type="l", col="red", xlab="Interval", ylab="Number of steps" )

```
