---
title: "Coursera 'Reproductible Research'- Assignment #1"
author: "Abdel612"
date: "May, 11th of 2015"
---

This is the R Markdown document as required by the Coursera "Reproductible Research" course's assignement #1.  



# Helper Functions
Let's start with some helper functions that allow us to uncrompress the provided data file , reading and filling the dataset up with measured values, and ending with a function computing the total number of step done in a day..

### Uncompress the provided data file (unziping the .zip file)
This function does not return anything, but makes the job to unzip the archive given by parameter *fileName*:
```{r, echo=TRUE}
#
# Unzip data archive file. Default file name="activity.zip"
#
uncompressDataArchive = function(fileName="activity.zip"){ 
    unzip(fileName, files = NULL, list = FALSE, overwrite = TRUE,
          junkpaths = FALSE, exdir = ".", unzip = "internal",
          setTimes = FALSE)
}
```

### Read the data file and load values in the returned data frame
Load any CSV formatted file given by parameter *fromFileName*

Returns a data frame corresponding to the CSV file
```{r, echo=TRUE}
#
# Load dataset according to a file name which defaulted to "activity.csv"
#
loadDataset = function(fromFileName="activity.csv"){
  dataSet=read.csv(fromFileName)
  return(dataSet)
}
```

### Compute the total number of steps by date. 
Returns a data frame made of 2 variables: 

- date : POSIX : the considered date
- steps: numeric : Total number of steps done the day date
```{r, echo=TRUE}
#
# Compute mean steps by date
#
# Return a new data.frame with 2 columns:
#     - date : POSIX Date in YYYY-MM-DD format
#     - steps: numeric : Total number of steps in the day "date"
getNumberOfStepsByDate = function(dataSet){
  totalStepPerDay = tapply(dataSet$steps, dataSet$date, sum, na.rm = TRUE)
  theDate = names(totalStepPerDay)
  theTotal = as.numeric(totalStepPerDay)
  result = data.frame(date=strptime(theDate, format="%Y-%m-%d"), steps=theTotal)
  return( result )
}
```



Y
ou can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
